---
title: "Week 3 Project"
author: "jrfoster"
date: "February 9, 2017"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r setupData}
suppressMessages(library(plotly))
suppressMessages(library(dplyr))

# Download the data if we don't already have it
if (!file.exists("pml-training.csv")) {
  download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                destfile = "pml-training.csv")
}

# Read the saved dataframe, otherwise, load the csv and save it as a dataframe
if (!file.exists("pmlTraining.rds")) {
  pmlRaw <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"))
  saveRDS(pmlRaw, file = "pmlTraining.rds")
} else {
  pmlRaw <- readRDS("pmlTraining.rds")
}

# Use dplyr to select the columns we will need from our plot which are only those
# related to the arm magnet and the classifier
pmlPlotting <- pmlRaw %>%
  select(magnet_arm_x, magnet_arm_y, magnet_arm_z, user_name, classe) %>%
  filter(classe == "A")
```

## Introduction

For week 3, we are asked to put together a presentation containing an interactive plot created with Plotly

This presentation contains one interesting visualization of some of the data used in Practical Machine Learning, specifically a 3-D scatterplot of one of the sensors used to map a bicep curl.

**One note to the reviewer - please view this on IE, Chrome or Safari as the plot will not render at all on Firefox.**


Enjoy!

## Visualization of the Arm Magnet Sensor

```{r plotData}
# Create the plot and display it
p <- plot_ly(pmlPlotting, x=~magnet_arm_x, y=~magnet_arm_y, z=~magnet_arm_z, color=~user_name) %>%
  add_markers()
p
```

## Code

```{r code, echo=TRUE, eval=FALSE}
suppressMessages(library(plotly))
suppressMessages(library(dplyr))

# Download the data if we don't already have it
if (!file.exists("pml-training.csv")) {
  download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                destfile = "pml-training.csv")
}

# Read the saved dataframe, otherwise, load the csv and save it as a dataframe
if (!file.exists("pmlTraining.rds")) {
  pmlRaw <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"))
  saveRDS(pmlRaw, file = "pmlTraining.rds")
} else {
  pmlRaw <- readRDS("pmlTraining.rds")
}

# Use dplyr to select the columns we will need from our plot which are only those
# related to the arm magnet and the classifier
pmlPlotting <- pmlRaw %>%
  select(magnet_arm_x, magnet_arm_y, magnet_arm_z, user_name, classe) %>%
  filter(classe == "A")

# Create the plot and display it
p <- plot_ly(pmlPlotting, x=~magnet_arm_x, y=~magnet_arm_y, z=~magnet_arm_z, color=~user_name) %>%
  add_markers()
p
```